To be FAIR: Theory Specification Needs an Update

### 2025-06-21 action_letter.Rmd
update 2025-06-21 06:45:52.895789

[x] Headings should not be numbered.
[x] Tables and Figures should be uploaded separately and not positioned in the text (although it is fine to indicate where they should be inserted).
[] Figures should be uploaded in their native format (the program they were created in) at 300 dpi. Acceptable figure file formats include MS Office files (DOC, PPT, XLS), Adobe Illustrator (AI), JPG, EPS, and PDF.
[x] References should be done with hanging indents using APA style. 
[] If you want to add the doi that is good.  But do not put in the search criteria for retrieving a document that has been published in a book or journal. 
```

In response to your request, we have: 
- uploaded tables separately: tab_meehl.csv
- uploaded figures separately: ec_visio.pdf, git_branches.png, git_compare_short.png

@@

We understand that striking the balance between the conceptual and technical details is the very challenge of this paper. In the revision we have tried to explain technical details in even simpler language.
The most important change we have made, as requested by Reviewer #2, is to move the tutorial parts of the paper into package vignettes.
We still leave a conceptual overview of the tutorial in simpler language because we do want readers of the paper to come away with some idea of what making a theory FAIR entails. Note that these general principles are neither tied to the specific R-pacakge not to our specific implementation. Importantly, they could be implemented using an entirely different infrastructure.

@@

We appreciate the Editor’s meta-scientific comparison of the emergence of conventions in human genetics to the (current wave of) more top-down nature of Open Science reforms in Psychology, which by-and-large also originated in grass-roots initiatives.

@@

To address this comment, in the revision, we repeatedly refer to the *potential* impact of FAIR theory, thus making it clear that we do not assume that this impact will be automatic or guaranteed. Furthermore, we acknowledge in the Limitations section that empirical evidence for these potential benefits remains to be evaluated:

@@

We thank the Reviewer for their thoughtful and encouraging review.
We appreciate the Reviewer's recognition of the relevance and potential impact of FAIR theory --- not only within psychology, but also in related disciplines.
We have sought to address all comments, and feel that doing so has improved the quality of the manuscript.

@@

As suggested, we now briefly mention the R package in the abstract:

@@

**DO THIS**

@@

We have done as requested, and additionally, include a sentence in the first paragraph to pre-emptively address readers' concerns that FAIR theory may only be relevant for those who have adopted a specific philosophical paradigm:

@@

We added a brief description and a reference as suggested:

`r get_revision("phonoloop")`

@@

We respectfully disagree; Open Data - one example of modular publishing - has increased dramatically in recent decades. In Psychology the percentage of articles that share data openly has increased from ~7% in 2000–2009 to ~50% in 2010-2019. In other disciplines, the increase has been even larger (up to ~80%).
Given this, we don't see a reason to note this as a limitation, but we have added a reference that indicates that data sharing has become widely adopted in Psychology:

@@

> We are particularly indebted to one of our reviewers, Dr. Eric Turkheimer, whose thoughtful reflections on FAIR theory were incorporated into the Discussion.

>>>>

[x] Headings should not be numbered.
[x] Tables and Figures should be uploaded separately and not positioned in the text (although it is fine to indicate where they should be inserted).
[] Figures should be uploaded in their native format (the program they were created in) at 300 dpi. Acceptable figure file formats include MS Office files (DOC, PPT, XLS), Adobe Illustrator (AI), JPG, EPS, and PDF.
[x] References should be done with hanging indents using APA style. 
[] If you want to add the doi that is good.  But do not put in the search criteria for retrieving a document that has been published in a book or journal. WHAT??
```

We have 
Upload tables separately: tab_meehl.csv
Upload figures separately: ec_visio.pdf, git_branches.png, git_compare_short.png

@@

We understand that striking the balance between the conceptual and technical details is the very challenge of this paper. In the revision we have tried to explain technical details in even simpler language.
The most important change we have made, as requested by Reviewer #2, is to move the tutorial parts of the paper into package vignettes.
We still leave a conceptual overview of the tutorial in simpler language because we do want readers of the paper to come away with some idea of what making a theory FAIR entails, and because these general principles are not tied to the R-pacakge and our specific implementation (they could be implemented using other infrastructures).

@@

We appreciate the Editor’s meta-scientific comparison of the emergence of conventions in human genetics to the (current wave of) more top-down nature of open science reforms in psychology, which by-and-large also originated in grass-roots initiatives.

@@

To address this comment, in the Revision, we repeatedly refer to the *potential* impact of FAIR theory, thus making it clear that we do not assume that this impact will be automatic nor guaranteed. Furthermore, we acknowledge in the Limitations section that empirical evidence for these potential benefits remains to be evaluated:

@@

We thank the Reviewer for their thoughtful and encouraging review.
We appreciate the Reviewer's recognition of the relevance and potential impact of FAIR theory—not only within psychology, but also in related disciplines.
We have sought to address all comments, and feel that doing so has improved the quality of the manuscript.

@@

As suggested, we now briefly mention the R-package in the abstract:

@@

We have rewritten this section so that it is clear that the two phases were not part of the original empirical cycle, but introduced by Wagenmakers (1b).

We have also made the two phases visually more distinct by adding a different background to each in 1b.

Finally, we now briefly introduce 1c as well:

`r get_revision("introduce1c")`

@@

We have done as requested, and additionally, include a sentence in the first paragraph to pre-emptively address readers' concerns that FAIR theory may only be relevant for those who have adopted a specific paradigm:

@@

**DO THIS, ANDREAS?**

@@

We respectfully disagree; open data - one example of modular publishing - has increased dramatically in recent decades, in psychology the percentage of articles that share data has increased from ~7% in 2000–2009 to ~50% in 2010-2019. In other disciplines, the increase has been even larger (up to ~80%).
Given this, we don't see a reason to note this as a limitation, but we have added a reference that indicates that data sharing has become widely adopted in psychology:

@@

> We are particularly endebted to our Reviewer, Dr. Eric Turkheimer, whose thoughtful reflections on FAIR theory were incorporated into the Discussion.

### 2025-06-21 test.Rmd
update 2025-06-21 06:45:52.895789

According to the *empirical cycle* [@degrootMethodologyFoundationsInference1969],
a meta-theoretical model of cumulative knowledge acquisition,
research ideally follows a cyclical process (Figure \@ref(fig:figecs)a).
Naturalistic observations or patterns identified in data give rise to preliminary hypotheses via induction.
Then, deduction is used to derive testable predictions from these hypotheses.
Finally, the outcomes of tests are evaluated with regard to their implications for theory.
Wagenmakers and colleagues divided this cycle into two phases [-@wagenmakersCreativityVerificationCyclePsychological2018, Figure \@ref(fig:figecs)b].
In one half of the cycle, which they labeled the "Context of Justification",
hypotheses derived from theory are tested on data.
In the other half of the cycle (the "Context of Discovery"),
patterns observed in data are generalized to theoretical principles.
<!-- Hier iets over mijn interpretatie -->
In this model, theories are the vehicle of scientists' understanding of phenomena.

@@

```{r figecs, eval = TRUE, fig.cap='Three implementations of the "empirical cycle" [@degrootMethodologyFoundationsInference1969].'}
# rsvg::rsvg_pdf("empirical_cycle2.svg", "ec_visio.pdf")
knitr::include_graphics("ec_visio.pdf")
```

@@

In implementing these different versions,
the first author realized that their work is guided by yet another different interpretation of the empirical cycle.
First, in De Groot's formulation, each stage describes a process.
This invites the question of what the concrete outcomes of these processes are.
For example, note the label "knowledge accumulation" in Wagenmakers' version.
This label invites the question of exactly *how* knowledge accumulates upon evaluation of a prior experiment.
What actually changes when going through the cycle, except the scholar's mind?
To address this point, the nodes in Van Lissa's specification (2025) refer to specific deliverables, and the edges connecting the nodes describe processes acting upon those deliverables (Figure \@ref(fig:figecs)c).
In this specification, `theory` can be understood as FAIR theory.

A second distinction is that the processes of induction and deduction are perhaps not as neatly confined to specific phases as De Groot proposed.
Theory testing, as takes place in the "context of justification", can be said to involve *mostly* deductive reasoning.
Theory development and amendment, as takes place in the "context of discovery", involves mostly inductive reasoning^[Here, "induction" is defined forming general theories based on specific observations. Others have used the term "abduction" to describe "inference to the best explanation" [@peirceCollectedPapersCharles1960]. For present purposes, the terms are interchangeable.].
However, deriving hypotheses from theory is not *entirely* deductive
as auxiliary assumptions must often be made to account for remaining ambiguities in theory, which involves induction.
A rudimentary example is assuming equal variances across groups when testing a mean difference between groups, because groups often have equal variances (induction from prior knowledge), or because a Levene's test on the dataset at hand is non-significant (induction from a specific observation to the population).
Similarly, if we accept that observation is theory-laden, then it too involves induction [@brewerTheoryLadennessObservationTheoryLadenness2001].
Similarly, if the testing procedure is not explicitly defined before seeing the data, it incurs some inductive bias as well [@peikertTransparencyOpenScience2023].
These alterations result in the following implementation of the empirical cycle:

```
theory;
prediction;
data;
test;
results;

theory -> prediction [label="deduction"];
prediction -> test [label = "implement inferential procedure"];
data -> results;
test -> results [label = "apply to data"];
results -> theory [label="interpretation and generalization"];
```

>>>>

According to the *empirical cycle* [@degrootMethodologyFoundationsInference1969],
a meta-theory of cumulative knowledge acquisition,
research ideally follows a cyclical process (Figure \@ref(fig:figecs)a).
Naturalistic observations or patterns identified in data give rise to preliminary hypotheses via induction.
Then, deduction is used to derive testable predictions from these hypotheses.
Finally, the outcomes of tests are evaluated with regard to their implications for theory.
Wagenmakers and colleagues divided this cycle into two phases [-@wagenmakersCreativityVerificationCyclePsychological2018, Figure \@ref(fig:figecs)b].
One phase is the "Context of Justification",
in which hypotheses derived from theory are tested on data.
The other phase is the "Context of Discovery",
in which patterns observed in data are generalized to theoretical principles.
<span id="introduce1c">In these interpretations of the empirical cycle, each stage describes a process.
This invites the question what the concrete outcomes of these processes are.
How does knowledge accumulate when iterating through the cycle; what actually changes, other than the scholar's mind?
To address these questions, the lead author specified a revised version of the empirical cycle (see "Changing a Theory")
in which the nodes refer to specific deliverables,
and the edges describe processes acting upon those deliverables (Figure \@ref(fig:figecs)c).
In this specification, `theory` refers to FAIR theory,
`hypothesis` to machine-readable hypotheses [@lakensImprovingTransparencyFalsifiability2021],
`test` refers to a preregistered inferential procedure [Preregistration-As-Code, @peikertReproducibleResearchTutorial2021],
data refers to FAIR data,
and results refer to (preprint) papers, ideally supplemented with a comprehensive reproducible research archive [@vanlissaWORCSWorkflowOpen2021].
In this model, theories are the vehicle of scientists' understanding of phenomena;
they are what changes when iterating through the cycle.</span>

@@

```{r figecs, eval = TRUE, fig.cap='Three implementations of the "empirical cycle" [@degrootMethodologyFoundationsInference1969].'}
# rsvg::rsvg_pdf("empirical_cycle3.svg", "ec_visio.pdf")
knitr::include_graphics("ec_visio.pdf")
```

@@

The present work is guided by an interpretation of the empirical cycle in which nodes refer to specific FAIR digital objects (FAIR theory, machine-readable hypotheses, FAIR data, et cetera),
and edges connecting the nodes refer to processes acting on (X-operating with) those digital objects (Figure \@ref(fig:figecs)c).
This implementation addresses an important ambiguity of prior versions of the empirical cycle.
De Groot describes how the phase of evaluation involves "formation of new [theories]";
evaluating results from one iteration of the cycle thus become observations (Phase 1) that initiate a new iteration.
De Groot stops short of detailing the process of theory construction or revision,
and in Wagenmakers' version, the term "knowledge accumulation" also remains unspecified.
Thus in both prior versions, it is unclear *how* knowledge accumulates in the evaluation phase.
In Van Lissa's specification (2025), `theory` is understood as FAIR theory,
which has clear procedures for making changes and connecting these changes to empirical evidence.

A second departure from De Groot is that the processes of induction and deduction are intentionally abandoned as distinct phases.
Theory testing, as takes place in the "context of justification", can be said to involve *mostly* deductive reasoning.
Theory development and amendment, as takes place in the "context of discovery", involves mostly inductive reasoning^[Here, "induction" is defined forming general theories based on specific observations. Others have used the term "abduction" to describe "inference to the best explanation" [@peirceCollectedPapersCharles1960]. For present purposes, the terms are interchangeable.].
However, deriving hypotheses from theory is not *entirely* deductive
as auxiliary assumptions must often be made to account for remaining ambiguities in theory, which involves induction [@peikertTransparencyOpenScience2023].
A common example is assuming equal variances across groups when testing a hypothesis about a mean difference between groups.
One might justify this assumption because groups often have equal variances (induction from prior knowledge), or because a Levene's test on the dataset at hand is non-significant (induction from a specific observation to the population).
Additionally, if we accept that observation is theory-laden, then it too involves induction [@brewerTheoryLadennessObservationTheoryLadenness2001].
These alterations result in the following implementation of the empirical cycle:

```
theory [label="fair theory"];
prediction [label="preregistration"];
data [label="fair data"];
test [label="reproducible analysis pipeline"];
results [label="preprint, reproducible research archive"];

theory -> prediction [label="derive hypothesis, refer to specific fair theory"];
prediction -> test [label = "implement inferential procedure as code"];
data -> results [label="apply preregistered analysis to data to obtain results"];
test -> results [label = "write preprint, create research archive"];
results -> theory [label="interpret findings and modify fair theory"];
```
